---
title: "Spatial aggregation confounds null model-based inference of species association networks"
author: "Andrew J. Rominger"
format:
  pdf:
    keep-tex: true
    include-in-header: 
      text: |
        \usepackage{xr}
        \externaldocument{ab_supp}
        \newcommand{\sfig}[1]{Supplementary Figure~\ref{#1}}
        \newcommand{\ssec}[1]{Supplementary Section~\ref{#1}}
        \usepackage{lineno}
csl: ecology.csl
bibliography: SpatAggNet.bib
---


```{r}
#| label: setup
#| include: false

library(knitr)
library(dplyr)
library(ggplot2)

opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# helper function to load cache from supp
load_supp <- function(cname) {
    f <- sprintf("%s.*RData", cname) |> 
        grep(list.files("SpatAggNet_supp_cache/pdf", full.names = TRUE), 
              value = TRUE) |> 
        gsub("\\.RData", "", x = _)
    
    o <- lazyLoad(f, envir = globalenv())
    
    invisible(o)
}

```


```{r}
#| label: supp
#| eval: false
#| include: false
#| cache: true
#| cache.extra: !expr file.info("SpatAggNet_supp.qmd")$mtime

# render the supplement so can be referenced and used in main

quarto::quarto_render("SpatAggNet_supp.qmd", quiet = FALSE)
```

```{r}
#| label: supp-tex-render
#| eval: false
#| include: false

# need .aux file for corss referencing
tinytex::xelatex("SpatAggNet_supp.tex", clean = FALSE)
```





\linenumbers

## Abstract 




## Introduction

Species interactions are fundamental to the evolutionary origins of biodiversity and the maintenance of that diversity in ecological assemblages [@hutchinson1961; @holt1994; @chesson2000; @proulx2005; @bascompte2007; @levine2009; @allesina2012; @johnson2019]. Network theory has come to be a central paradigm in ecology and evolution [@may1972; @williams2000; @dunne2002; @proulx2005; @bascompte2007], and so, not surprisingly, these interactions are often cast as ecological networks where species are nodes (or vertices) and network links (or edges) connect species that interact.

Abundance or presence/absence data are among the most prevalent and accessible ecological data, and have consequentially been frequently leveraged [@morueta2016; @calatayud2019] in an attempt to infer species interactions (or more accurately associations), beginning at least in the work of @diamond1975. The specific data used for constructing association networks are often site by species matrices where each cell in the matrix records the abundance (or presence/absence) of a given species at a given site (i.e. sampling location). For simplicity I will refer to both abundance and presence/absence data as simply "abundance data" and focus on true abundance data with the assumed implication that methodological limitations in the case of true abundances will be even more confounding for presence/absence scenarios [@ulrich2010].

Two salient insights from decades of debate about using abundance data to infer ecological processes are: 1) interactions in the strict sense cannot be inferred, but *species associations* might be recoverable and these *associations* are assumed to map onto a combination of species-species and species-environment interactions [@morueta2016; @calatayud2019]; and 2) robust null models are necessary to identify which patterns in the data are likely to arise by chance and which are likely to carry biological signal [@ulrich2010]. 

Null models seek to generate plausible data, and patterns in data, while stripping away the actual relationships between species in the data [@ulrich2010]. Most often a null model is achieved by repeatedly permuting the original data such that any real associations are removed because which species co-occur has been randomized by permutation [@ulrich2010]. The collection of all such permutations represents a null distribution. These permutation approaches also seek to preserve key aspects of the original data such as maintaining the observed abundances of each species and/or maintaining the observed total number of individuals at each site [@ladau2017]. These constraints are imposed in an attempt to control the rate of incorrect inference of the presence or absence of network links [@ladau2008; @ladau2017]. Other methods for inferring association networks do exist, especially model-based inference applied in microbial ecology [e.g. @zhou2010; @kurtz2015], but are less frequently employed in community ecology [but see, e.g., @harris2016; @losapio2021].

Here I raise a previously unreported technical issue with the approach of inferring association networks from abundance data using permutational null models. The issue arises from the near universal spatial aggregation of individuals within species [@legendre1993; @condit2000; @lennon2000; @dale2002; @mcgill2003; @engen2008; @harte2011; @zillio2010; @connolly2017]. Using simulation experiments and analysis of real data [from @calatayud2019], I show that intra-specific spatial aggregation (ISSA) causes exceedingly high type I error rates in permutational null model approaches to inferring association networks, i.e., incorrectly detecting associations when there are none. ISSA in itself is not evidence of species association---it can arise from a large number of processes including simple demographic drift [@kendall1949; @hubbell2001] in which species do not meaningfully interact with each other or shared environments. 

In the absence of ISSA, abundances of a species across sites should follow a Poisson distribution [@diggle]. Thus ISSA is often quantified as the deviation of observed data from a Poisson distribution [@diggle]. Conversely, the negative binomial distribution can model spatially aggregated intra-specific abundances across sites [@bliss1953; @taylor1961; @engen2008; @connolly2017]. Critically, the negative binomial arises very naturally from ecological processes as the distribution of population sizes in a birth-death-immigration model [@kendall1949]. If the abundance of a species at a site is determined by local birth and death with supplemental individuals arriving by immigration from a source pool, then the probability distribution of repeatedly sampled abundances will follow a negative binomial distribution [@kendall1949]. In this way, simple demographic drift (birth, death, and immigration that are purely stochastic) is expected to produce intra-specific abundances that are negative binomial-distributed and spatially aggregated. Thus ISSA is not, by itself, evidence of species association.

The technical issue raised here is specific to the widely used family of methods that infer association networks by comparing an observed test statistic to a null distribution generated by Monte Carlo permutation of the site-by-species matrix. I do not test whether the same issue affects methods that do not rely on this style of null model, most notably joint species distribution models, which explicitly condition on measured or latent environmental covariates and site-level random effects [e.g. @zhou2010], or the compositional correlation methods common in microbial ecology [e.g. @kurtz2015]. It is not obvious that the mechanism described here transfers directly to these other methods, and I do not want to imply, by omission, that it does. However, any effects of ISSA on joint species distribution modeling and compositional correlation methods should be assessed. Here I focus solely permutational null models because of their foundational status in the field [@cite] and their continued use in inferring association networks.

<!-- The present work complements a recent uptick in critical and cautionary assessments of approaches seeking to infer ecological network from abundance data [@freilich2018; @barner2018; @carr2019; @blanchet2020]. The combined message of these contributions is that abundance-derived inferences of species associations should be viewed cautiously until other data can be brought to bear on the actuality of real associations. -->

## Methods

### Reproducibility and data

A custom R package, *SpatAggNet* accompanies this paper and can be used to reproduce all results. ssec{sec-setup} details how to access and install *SpatAggNet*. All analysis scripts and data analyzed are available via the package.

I use a data set of site-by-species abundance matrices compiled by @calatayud2019. The data are included with the *SpatAggNet* package as `abund_mats` and comprise `|r length(abund_mats)` site-by-species matrices. These compiled data are ideal for investigating the impact of ISSA on association network inference because they were used by @calatayud2019 for the very purpose of network inference and also because they represent a spatially and taxonomically extensive collection of quality-controlled data.


### Quantifying spatial aggregation

The spatial species abundance distribution (SSAD) is the distribution of a single species' replicate abundances across sites. I fit both negative binomial and Poisson distributions to the SSADs of each species in each site-by-species matrix using maximum likelihood. The negative binomial distribution represents ISSA.

The `nb_fit()` function from *SpatAggNet* fits both distributions to every species in each site-by-species matrix and returns, for each species from each matrix: 

- the negative binomial dispersion parameter (`size`, i.e. $k$) and mean (`mu`)
- the log-likelihoods of negative binomial and Poisson models
- their AIC difference (`delta_AIC`), 
- a $\chi^2$-distributed goodness-of-fit statistic for the negative binomial [`zsq`; @etienne2007]
- the species' total abundance and occupancy.

The negative binomial likelihood function does not have a finite maximum when the mean of a species' abundances across sites is greater than or equal to their variance. A non-finite likelihood is not necessarily evidence against the negative binomial as the true generating distribution, as @fig-good-mle shows, samples drawn from a negative binomial distribution have a non-0 probability of producing an estimated mean greater than or equal to the estimated variance. This occurs disproportionately for rare species and for studies with few sites (@fig-inf-k), which is not entirely unexpected that rare species and/or matrices with less sampling will have poorly behaved likelihood functions. 

```{r}
#| label: fig-good-mle
#| cache: true
#| fig-width: 5
#| fig-height: 3
#| fig-cap: |
#|   The relationship between total species abundance and the probability
#|   that the negative binomial likelihood function is solvable, for
#|   data truly generated by a negative binomial distribution. Lines
#|   correspond to different negative binomial dispersion parameters and
#|   numbers of spatial replicates.

fig_good_mle
```

I therefore assess the prevalence of the negative binomial SSAD, and thus ISSA, in two complementary ways: 1) where the likelihood is defined, I compare AIC support for the negative binomial versus the Poisson; and 2) for all species, regardless of whether the likelihood is defined, I evaluate goodness of fit of the negative binomial using the square-transformed $z$ statistic of @etienne2007, which is $\chi^2$-distributed with one degree of freedom under the null hypothesis that the data are negative binomial [@romMeteR].

I also fit a log-log linear model relating each species' dispersion parameter $k$ to its relative abundance, which I later use to simulate realistic SSAD shapes for species of any given abundance.

### Analysis of empirical data

- purpose: show the issue and that nb is the reason
- data from rareplusminus
- get sad and ssad from those data
    - pois
    - nb
    - model selection of pois and nb
- look at network stats for real data versus data simulated from sad and ssad
    - pois: doesn't look like real data
    - nb: looks close to real data
- report on rareplusminus stuff

### Simulation experiment

- purpose: show the issue persists across different methods
- start with data simulated from sad and ssad
    - only need to do nb
- make networks based on diff perm methods and diff distance metrics


### Behavior of negative binomial random variables subjected to permutation 

- show that k is distorted 


## Results

### Analysis of empirical data

### Simulation experiment

## Discussion

## References
